upwell_IDX

### 2019-09-04 upwell_IDX.Rmd
Updated

```{r}
# wind_1 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind1/wind1.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub", "speed", "dir"))
# 
# wind_2 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind2/wind2.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub" ,"speed", "dir"))
# 
# wind_3 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind3/wind3.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub" ,"speed", "dir"))
# 
# # Slecting the important columns for each of the datasets
# wind_fix <- function(df){
# wind <- df %>%
#   select(station_name, date, hour, dir, speed) %>%  #column names
#   mutate(date = as.Date(as.character(date)),
#          hour = as.numeric(as.character(hour)),
#          dir = as.numeric(as.character(dir)),
#          speed = as.numeric(as.character(speed)))
# }
# 
# wind_1 <- wind_fix(df = wind_1)
# wind_2 <- wind_fix(df = wind_2)
# wind_3 <- wind_fix(df = wind_3)
# 
# ## Renaming the sites within the wind datasets to match the name of the sites at which seawater temperature was collected
# ## The wind data was obtained from the SAWS and the wind stations used were the closes stations to which temperature was collected
# 
# renaming_sites_1 <- function(df) {
#   sites <- df %>%
#       mutate(temp_sites = case_when(station_name == "CAPE TOWN TABLE BAY" ~ "Sea Point",
#                                     station_name == "CAPE TOWN - ROYAL YACHT CLUB" ~ "Sea Point",
#                                     station_name  == "PORT NOLLOTH" ~"Port Nolloth",
#                                     station_name  == "CAPE TOWN SLANGKOP" ~ "Hout Bay",
#                                     station_name  == "LAMBERTSBAAI NORTIER" ~ "Lamberts Bay",
#                                     station_name  == "LANGEBAANWEG AWS" ~ "Saldanha Bay"))
#   return(sites)
# }
# 
# 
# wind_sitesmatched_1 <-  renaming_sites_1(df = wind_1)
# wind_sitesmatched_2 <-  renaming_sites_1(df = wind_2)
# wind_sitesmatched_3 <-  renaming_sites_1(df = wind_3)
# wind_data <- rbind(wind_sitesmatched_3,wind_sitesmatched_2,wind_sitesmatched_1)
# # save(wind_data, file = "Data/wind_data.RData")
# 
# load("Data/wind_data.RData")
# selected_sites <- c("Port Nolloth", "Lamberts Bay", "Sea Point", "Saldanha Bay", "Hout Bay")
# 
# wind_daily <- wind_data %>%
#   # ungroup() %>%
#   # mutate(date = as.Date(date))%>%
#   dplyr::group_by(temp_sites, date) %>%
#   filter(temp_sites %in% selected_sites) %>%
#   dplyr::summarise(dir_circ = round(mean.circular(circular(dir, units = "degrees")),2),
#                    mean_speed = round(mean(speed),2))
# #save(wind_daily, file = "Data/wind_daily.RData")
```

@@

```{r}
south_africa_away_wide <- coastR::transects(site_list_sub, spread = 30)
world_map <- ggplot() + 
  borders(fill = "grey40", colour = "black")
# Create titles
titles <- c("Shore-normal", "Islands")
# Plotting function
plot_sites <- function(site_list, buffer, title_choice, dist){

@@

```{r}
UI_angle <- wind_daily %>% 
  dplyr::rename(site = sites) %>% 
  left_join(south_africa_away_wide[,c(2,21)], by = "site") %>% 
  dplyr::rename(coast_angle = heading) %>% 
  mutate(ui.saws = mean_speed * (cos(dir_circ - coast_angle))) %>% # Calculated perpendicular angle for each of the sites. Not necessary
  drop_na %>% 
  dplyr::rename(t = date)

# save(UI_angle, file = "Data/UI_angle.RData")

# index <- wind_UI %>%
#   group_by(site) %>%
#   select("date", "ui.saws") %>%
#   mutate(saws.condition = ifelse(ui.saws < 12.5, "upwelling"))

# wind_UI <- wind_daily %>% 
#   mutate(ui.saws = mean_speed * (cos(dir_circ - 160))) %>% # 160 is angle perpendicular to the west coast of SA (See paper Fielding and Davis pg 182)
#   drop_na
# Results obtained from UI_angle (Using the perpendicular angle for each site) and the results of wind_UI at -160 (angle of the west coast) appears to be the same
```

# Exceedence function heatwaveR - This is done to determine the consecutive number of days at or above what the UI value is meant to be

```{r}
# Loading the insitu temperature data along the wc
load("Data/SACTN_US.RData")
# exceedance <-
#   function(data,
#            x = t,
#            y = temp,
#            threshold,
#            below = FALSE,
#            minDuration = 5,
#            joinAcrossGaps = TRUE,
#            maxGap = 2,
#            maxPadLength = 3)
# Upwelling temperature threshold 12.4 / 30th percentile
SACTN_upwell <- UI_angle %>% 
  dplyr::rename(temp = ui.saws) %>%
  group_by(site) %>%
  # mutate(min_t = min(t), 
  #        max_t = max(t)) %>% 
  nest() %>% 
  mutate(clim = purrr::map(data, ts2clm, climatologyPeriod = c("2000-01-01", "2005-12-31")),
         # NB: A threshold of 3 appeared to be far to strict
         exceed = purrr::map(clim, exceedance, minDuration = 1, threshold = 1)) %>%  #Upwelling cannot be descrbed as an event. Upwelling can last for a few hours. Given that we have daily data, upwelling events minimum duration here will be 1day
  # mutate() %>% 
  select(-data, -clim) %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>%
  unnest() %>% 
  dplyr::rename(ui.saws = temp) %>% 
  select(site, t, ui.saws, exceedance)

# Calculate quantiles of upwelling index
SACTN_upwell_quantiles <- SACTN_upwell %>%
  filter(ui.saws >= 0) %>% 
  group_by(site) %>% 
  summarize(quant_10 = quantile(ui.saws, probs = 0.10, na.rm = TRUE),
            quant_25 = quantile(ui.saws, probs = 0.25, na.rm = TRUE),
            quant_50 = quantile(ui.saws, probs = 0.50, na.rm = TRUE),
            quant_75 = quantile(ui.saws, probs = 0.75, na.rm = TRUE),
            quant_90 = quantile(ui.saws, probs = 0.90, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, digits = 2)

@@

detect_event_custom <- function(df){
  res <- detect_event(df, threshClim2 = df$exceedance, minDuration = 3, coldSpells = T)
  return(res)
}
# Calculate the upwelling event metrics
SACTN_upwell_base <- SACTN_US %>% 
  dplyr::rename(t = date) %>% 
  group_by(site) %>% 
  nest() %>% 
  # The climatology base period used here is up for debate...
  # The choice of the 30th percentile threshold also needs to be justified and sensitivty tested
  mutate(clim = purrr::map(data, ts2clm, pctile = 25, climatologyPeriod = c("1995-01-01", "2004-12-31"))) %>%
  select(-data) %>% 
  unnest() %>%
  left_join(SACTN_upwell, by = c("site", "t")) %>%
  filter(!is.na(exceedance)) %>%
  group_by(site) %>% 
  # mutate(thresh = mean(seas, na.rm = T)-sd(seas, na.rm = T)) %>% # Manually set threshold to the mean
  # mutate(thresh = 10) %>% # Manually set threshold to a static value. Doesn't work across all sites
  # mutate(thresh = quantile(temp, 0.3, na.rm = T)) %>% # Manually set threshold to a single quantile
  nest() %>% 
  mutate(exceed = purrr::map(data, detect_event_custom)) %>% 
  select(-data) #%>% 
  # unnest() %>% 
  # filter(row_number() %% 2 == 0) %>% # Select event summary metrics
  # filter(row_number() %% 2 == 1) %>% # Select daily values
  # unnest()

# Unpack the event metric reults
SACTN_upwell_events <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 0) %>%
  unnest()
# save(SACTN_upwell_events, file = "Data/SACTN_upwell_events.RData")

# Unpack the daily climatology results
SACTN_upwell_clims <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>% 
  unnest()
# save(SACTN_upwell_clims, file = "Data/SACTN_upwell_clims.RData")

# The above chunk of code appears to work as expected.
# What needs to be done now is that the thresholds for duration and UI strength need to be justified/decided uopn. 
    # Min duration for an upwelling event - 1day and threshold -1
# What also needs to be decided is if we are interested in any upwelling results throughout the year,
# or only during the upwelling season, whenever that may be for each site.
# I think we have demonstrated the technical capability needed to answer the question of whether or not upwelling
# is changing over time.
# Now we really need to focus on which of these parameters need to best be tweaked to answer that question.
```

@@

```{r}
# match_func <- function(df){
#   match <- SACTN_MCS_clims  %>% 
#   left_join(df, by = c("site", "t")) %>% 
#   na.trim()
#   return(match)
# }
# 
# index <- index %>% 
#   dplyr::rename(site = sites) %>% 
#   dplyr::rename(t = date)
# 
# match <- match_func(df = index)
# 
# upwell_MCS <- ggplot(data = match, aes(x = t, y = ui.saws)) +
#   geom_line() +
#   geom_point(data = filter(match, event == TRUE), colour = "navy") +
#   geom_rug(data = filter(match, saws.condition == "upwelling"), sides = "b") +
#   scale_x_date(expand = c(0, 0)) +
#   facet_wrap(~site, ncol = 1) +
#   theme_grey() +
#   labs(y = "Temperature (°C)", x = NULL)
# upwell_MCS
# Plot some of the Upwelling/MCS results
SP_upwelling <- SACTN_upwell_clims %>% 
  filter(site == "Sea Point", lubridate::year(t) %in% 2016:2018)
ggplot(SP_upwelling, aes(x = t, y = temp)) +
  geom_flame(aes(y = thresh, y2 = temp), fill = "navy", n = 3) +
  geom_point(data = filter(SP_upwelling, !is.na(event_no))) +
  geom_line(aes(y = thresh)) +
  geom_line(alpha = 0.7) +
  geom_rug(data = filter(SP_upwelling, event_no > 0), sides = "b")

# We can see in the figure above that the rug plot, which denotes when the wind driven upwelling was detected,
# occurrs much less frequently than simply using the 30th percentile threshold.
# The dots on the plot help to show the length of each wind driven upwelling event detected.
# This supports the use of this methodology.
# Up next is fitting linear models to the upwelling metric results to see what trends shake out.

# Linear Models
lm_coeff <- function(df){
  res <- lm(formula = val ~ date_peak, data = df)
  res_coeff <- round(as.numeric(res$coefficient[2]), 4)
}
# Changes in upwelling metrics
SACTN_upwell_lm <- SACTN_upwell_events %>% 
  select(-c(index_start:index_end)) %>% 
  gather(key = "var", value = "val", -c(site:date_end)) %>% 
  group_by(site, var) %>% 
  nest() %>% 
  mutate(slope = purrr::map(data, lm_coeff)) %>% 
  select(-data) %>% 
  unnest() %>% 
  # convert from daily to decadal values
  mutate(slope = slope * 365.25*10)
```

>>>>

```{r}
# wind_1 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind1/wind1.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub", "speed", "dir"))
# 
# wind_2 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind2/wind2.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub" ,"speed", "dir"))
# 
# wind_3 <- read.delim("Data/Wind_data/wind_data.txt(SAWS)/wind3/wind3.txt", na.strings = "",
#                      col.names = c("station_number", "station_name", "date", "hour", "sub" ,"speed", "dir"))
# 
# # Slecting the important columns for each of the datasets
# wind_fix <- function(df){
# wind <- df %>%
#   select(station_name, date, hour, dir, speed) %>%  #column names
#   mutate(date = as.Date(as.character(date)),
#          hour = as.numeric(as.character(hour)),
#          dir = as.numeric(as.character(dir)),
#          speed = as.numeric(as.character(speed)))
# }
# 
# wind_1 <- wind_fix(df = wind_1)
# wind_2 <- wind_fix(df = wind_2)
# wind_3 <- wind_fix(df = wind_3)
# 
# ## Renaming the sites within the wind datasets to match the name of the sites at which seawater temperature was collected
# ## The wind data was obtained from the SAWS and the wind stations used were the closes stations to which temperature was collected
# 
# renaming_sites_1 <- function(df) {
#   sites <- df %>%
#       mutate(temp_sites = case_when(station_name == "CAPE TOWN TABLE BAY" ~ "Sea Point",
#                                     station_name == "CAPE TOWN - ROYAL YACHT CLUB" ~ "Sea Point",
#                                     station_name  == "PORT NOLLOTH" ~"Port Nolloth",
#                                     station_name  == "CAPE TOWN SLANGKOP" ~ "Hout Bay",
#                                     station_name  == "LAMBERTSBAAI NORTIER" ~ "Lamberts Bay",
#                                     station_name  == "LANGEBAANWEG AWS" ~ "Saldanha Bay"))
#   return(sites)
# }
# 
# 
# wind_sitesmatched_1 <-  renaming_sites_1(df = wind_1)
# wind_sitesmatched_2 <-  renaming_sites_1(df = wind_2)
# wind_sitesmatched_3 <-  renaming_sites_1(df = wind_3)
# wind_data <- rbind(wind_sitesmatched_3,wind_sitesmatched_2,wind_sitesmatched_1)
# # save(wind_data, file = "Data/wind_data.RData")
# 
# load("Data/wind_data.RData")
# selected_sites <- c("Port Nolloth", "Lamberts Bay", "Sea Point", "Saldanha Bay", "Hout Bay")
# 
# wind_daily <- wind_data %>% 
#   # ungroup() %>%
#   # mutate(date = as.Date(date))%>% # Changing the date to date format 
#   dplyr::group_by(temp_sites, date) %>% 
#   filter(temp_sites %in% selected_sites) %>%  # Matching the sites to wind temp sites
#   dplyr::summarise(dir_circ = round(mean.circular(circular(dir, units = "degrees")),2), # Direction using the circular function
#                    mean_speed = round(mean(speed),2)) # Calculating the mean 
# #save(wind_daily, file = "Data/wind_daily.RData")
```

@@

```{r}
# Coast R- function that calculates the angle of the site perpendicular to the coastline
# Transect function in coastR package finds the angle away or along the coastline
south_africa_away_wide <- coastR::transects(site_list_sub, spread = 30)

# Plotting the map
world_map <- ggplot() + 
  borders(fill = "grey40", colour = "black")

# Create titles
titles <- c("Shore-normal", "Islands")
# Plotting function
plot_sites <- function(site_list, buffer, title_choice, dist){

@@

# Upwelling indeces using the above formula

```{r}
UI_angle <- wind_daily %>%  # Making refeerence to the wind daily datta created above
  dplyr::rename(site = sites) %>%  # Renaming sites to site
  left_join(south_africa_away_wide[,c(2,21)], by = "site") %>%  # Joining wind data to south africa wide data column 2 and 21. by site 
  dplyr::rename(coast_angle = heading) %>% # Renamingthe column heading to coast_angle
  mutate(ui.saws = mean_speed * (cos(dir_circ - coast_angle))) %>% # applying the ui formula
  drop_na %>% # removing the na values
  dplyr::rename(t = date) #renaming date to t

# save(UI_angle, file = "Data/UI_angle.RData")
# index <- wind_UI %>%
#   group_by(site) %>%
#   select("date", "ui.saws") %>%
#   mutate(saws.condition = ifelse(ui.saws < 12.5, "upwelling"))
# wind_UI <- wind_daily %>% 
#   mutate(ui.saws = mean_speed * (cos(dir_circ - 160))) %>% # 160 is angle perpendicular to the west coast of SA (See paper Fielding and Davis pg 182)
#   drop_na
# Results obtained from UI_angle (Using the perpendicular angle for each site) and the results of wind_UI at -160 (angle of the west coast) appears to be the same
```

# Exceedence function heatwaveR - This is done to determine the consecutive number of days at or above what the UI value is meant to be   # Detect consecutive days in exceedance of a given threshold.

```{r}
# Loading the insitu temperature data along the wc
load("Data/SACTN_US.RData") # temperature data for all the sites with a 30year time series
# exceedance <-
#   function(data,
#            x = t,
#            y = temp,
#            threshold,
#            below = FALSE,
#            minDuration = 5,
#            joinAcrossGaps = TRUE,
#            maxGap = 2,
#            maxPadLength = 3)
# Upwelling temperature threshold 12.4 / 30th percentile
SACTN_upwell <- UI_angle %>% 
  dplyr::rename(temp = ui.saws) %>%
  group_by(site) %>%
  # mutate(min_t = min(t), 
  #        max_t = max(t)) %>% 
  nest() %>% # apply the following functions to all of the variables in te dataset
  mutate(clim = purrr::map(data, ts2clm, climatologyPeriod = c("2000-01-01", "2005-12-31")), # creating a column of climaatologies. Column will be named clim
         # NB: A threshold of 3 appeared to be far to strict
         # purr::map - apllies a function to each element of a vector
         exceed = purrr::map(clim, exceedance, minDuration = 1, threshold = 1)) %>%  #Upwelling cannot be descrbed as an event. Upwelling can last for a few hours. Given that we have daily data, upwelling events minimum duration here will be 1day
  # Detect consecutive days in exceedance of a given threshold.
  # mutate() %>% 
  select(-data, -clim) %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>%
  unnest() %>% # creates a column for each variables
  dplyr::rename(ui.saws = temp) %>% # rename upwelling index vale to temp so that it could work with the function
  select(site, t, ui.saws, exceedance) # selecting only these variables
# Calculate quantiles of upwelling index

# Static numbers are often rejected and so we decided to find a percentile value as these are often more likely to be approved

SACTN_upwell_quantiles <- SACTN_upwell %>% 
  filter(ui.saws >= 0) %>% # Upwelling occurs for all values above 0. Values below this is regarded as downwelling
  group_by(site) %>% 
  summarize(quant_10 = quantile(ui.saws, probs = 0.10, na.rm = TRUE),
            quant_25 = quantile(ui.saws, probs = 0.25, na.rm = TRUE),
            quant_50 = quantile(ui.saws, probs = 0.50, na.rm = TRUE),
            quant_75 = quantile(ui.saws, probs = 0.75, na.rm = TRUE),
            quant_90 = quantile(ui.saws, probs = 0.90, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, digits = 2)

@@

# Detect event: 
detect_event_custom <- function(df){
  res <- detect_event(df, threshClim2 = df$exceedance, minDuration = 3, coldSpells = T)
  return(res)
}
# Calculate the upwelling event metrics
SACTN_upwell_base <- SACTN_US %>% 
  dplyr::rename(t = date) %>% 
  group_by(site) %>% 
  nest() %>% 
  # The climatology base period used here is up for debate...
  # The choice of the 30th percentile threshold also needs to be justified and sensitivty tested
  mutate(clim = purrr::map(data, ts2clm, pctile = 25, climatologyPeriod = c("1995-01-01", "2004-12-31"))) %>%
  select(-data) %>% 
  unnest() %>%
  left_join(SACTN_upwell, by = c("site", "t")) %>%
  filter(!is.na(exceedance)) %>%
  group_by(site) %>% 
  # mutate(thresh = mean(seas, na.rm = T)-sd(seas, na.rm = T)) %>% # Manually set threshold to the mean
  # mutate(thresh = 10) %>% # Manually set threshold to a static value. Doesn't work across all sites
  # mutate(thresh = quantile(temp, 0.3, na.rm = T)) %>% # Manually set threshold to a single quantile
  nest() %>% 
  mutate(exceed = purrr::map(data, detect_event_custom)) %>% 
  select(-data) #%>% 
  # unnest() %>% 
  # filter(row_number() %% 2 == 0) %>% # Select event summary metrics
  # filter(row_number() %% 2 == 1) %>% # Select daily values
  # unnest()
# Unpack the event metric reults
SACTN_upwell_events <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 0) %>%
  unnest()
# save(SACTN_upwell_events, file = "Data/SACTN_upwell_events.RData")
# Unpack the daily climatology results
SACTN_upwell_clims <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>% 
  unnest()
# save(SACTN_upwell_clims, file = "Data/SACTN_upwell_clims.RData")
# The above chunk of code appears to work as expected.
# What needs to be done now is that the thresholds for duration and UI strength need to be justified/decided uopn. 
    # Min duration for an upwelling event - 1day and threshold -1
# What also needs to be decided is if we are interested in any upwelling results throughout the year,
# or only during the upwelling season, whenever that may be for each site.
# I think we have demonstrated the technical capability needed to answer the question of whether or not upwelling
# is changing over time.
# Now we really need to focus on which of these parameters need to best be tweaked to answer that question.
```

@@

```{r}
# match_func <- function(df){
#   match <- SACTN_MCS_clims  %>% 
#   left_join(df, by = c("site", "t")) %>% 
#   na.trim()
#   return(match)
# }
# 
# index <- index %>% 
#   dplyr::rename(site = sites) %>% 
#   dplyr::rename(t = date)
# 
# match <- match_func(df = index)
# 
# upwell_MCS <- ggplot(data = match, aes(x = t, y = ui.saws)) +
#   geom_line() +
#   geom_point(data = filter(match, event == TRUE), colour = "navy") +
#   geom_rug(data = filter(match, saws.condition == "upwelling"), sides = "b") +
#   scale_x_date(expand = c(0, 0)) +
#   facet_wrap(~site, ncol = 1) +
#   theme_grey() +
#   labs(y = "Temperature (°C)", x = NULL)
# upwell_MCS
# Plot some of the Upwelling/MCS results
SP_upwelling <- SACTN_upwell_clims %>% 
  filter(site == "Sea Point", lubridate::year(t) %in% 2016:2018)
plot1 <- ggplot(SP_upwelling, aes(x = t, y = temp)) +
  geom_flame(aes(y = thresh, y2 = temp), fill = "navy", n = 3) +
  geom_point(data = filter(SP_upwelling, !is.na(event_no))) +
  geom_line(aes(y = thresh)) +
  geom_line(alpha = 0.7) +
  geom_rug(data = filter(SP_upwelling, event_no > 0), sides = "b")
plot1

# The dots on the plot help to show the length of each wind driven upwelling event detected.
# This supports the use of this methodology.
# Up next is fitting linear models to the upwelling metric results to see what trends shake out.
# Linear Models
lm_coeff <- function(df){
  res <- lm(formula = val ~ date_peak, data = df)
  res_coeff <- round(as.numeric(res$coefficient[2]), 4)
}
# Changes in upwelling metrics
SACTN_upwell_lm <- SACTN_upwell_events %>% 
  select(-c(index_start:index_end)) %>% 
  gather(key = "var", value = "val", -c(site:date_end)) %>% 
  group_by(site, var) %>% 
  nest() %>% 
  mutate(slope = purrr::map(data, lm_coeff)) %>% 
  select(-data) %>% 
  unnest() %>% 
  # convert from daily to decadal values
  mutate(slope = slope * 365.25*10)
```
